---
title: "Lab 2"
author: "Jacob Jameson"
output: html_document
---


```{r, include=F,message=F}
library(knitr)
```

### Life Expectancy Estimation with the DEALE Method

Suppose that the 5-year survival probability for 60-year-old white women with disease Y is 60%, and the 10-year survival probability for 60-year-old white women with disease Z is 60%. The estimated life expectancy for a 60-year-old woman is 23.2 years, and the estimated life expectancy for a 55-year-old woman is 27.5 years.

(a) Using the DEALE method, calculate the life expectancy for a 55-year-old white woman with disease Y.

<span style="color:red">SOLUTION</span>

If we assume mortality from Y is constant by age, then we can “borrow” the disease- specific mortality from the 60-year-old white woman with disease Y:
Excess Mortality of Disease:

\[\mu_c = \mu_{ASR} + \mu_Y\]

We want to know $\mu_Y \implies \mu_Y = \mu_c - \mu_{ASR}$.

We are given the 5-year probability of survival with disease Y, and using the formula $S(t) = e^{-\mu t}$:

\[S(5) = e^{-5\mu_c} = 0.6\]

\[-5\mu_c = ln(0.6)\]

\[\mu_c = \frac{ln(0.6)}{-5}\]

```{r}
mu.c.60 <- log(0.6)/(-5)
print(mu.c.60)
```

We know that Life Expectancy at age 60 is 23.2 ($LE_{60} = 23.2$). We can use this to calculate $\mu_{ASR}$:

\[LE_{60} = 23.2 \implies \mu_{ASR} = \frac{1}{23.2}\]

```{r}
mu.asr.60 <- 1/23.2
print(mu.asr.60)
```

Now that we know $\mu_{ASR}$ and $\mu_{c}$, we can calculate $\mu_Y$:

```{r}
mu.y <- mu.c.60 - mu.asr.60
print(mu.y)
```

Now we can calculate $\mu_{ASR}$ for the 55 year old woman. We know that Life Expectancy at age 55 is 27.5 ($LE_{55} = 27.5$):

\[LE_{55} = 27.5 \implies \mu_{ASR} = \frac{1}{27.5}\]

```{r}
mu.asr.55 <- 1/27.5
print(mu.asr.55)
```

Now we can estimate life expectancy using DEALE, which states that:

\[LE = \frac{1}{\mu} = \frac{1}{\mu_{ASR} + \mu_Y}\]

For the 55 year old woman, assuming that $\mu_Y$ is constant between ages:

```{r}
LE.55 <- 1/(mu.asr.55 + mu.y)
print(LE.55)
```



(b) Using the DEALE method, calculate the life expectancy for 55-year-old white women with both disease Y and disease Z.

<span style="color:red">SOLUTION</span>


In this case, we need to solve for $\mu_Z$, therefore excess mortality of disease is:

\[\mu_c = \mu_{ASR} + \mu_Y + \mu_Z\]

We know that the 10-year survival probability for 60-year-old white women with disease Z is 60\%:

\[S(10) = e^{-10\mu_c} = 0.6\]

\[-10\mu_c = ln(0.6)\]

\[\mu_c = \frac{ln(0.6)}{-10}\]

```{r}
mu.c.60 <- log(0.6)/-10
print(mu.c.60)
```

We have already calculated $\mu_{ASR}$ for the 55-year old woman, and we assume $\mu_Y$ to be constant, so let us also assume $\mu_Z$ to be constant. 

Solving for $\mu_Z$:

\[\mu_Z = \mu_c - \mu_{ASR}\]

```{r}
mu.z <- mu.c.60 - mu.asr.60 
print(mu.z)
```

Now we can estimate life expectancy using DEALE, which states that:

\[LE = \frac{1}{\mu} = \frac{1}{\mu_{ASR} + \mu_Y + \mu_Z}\]

For the 55 year old woman, assuming that $\mu_Y$ and $\mu_Z$ is constant between ages:

```{r}
LE.55 <- 1/(mu.asr.55 + mu.y + mu.z)
print(LE.55)
```


(c) Consider a treatment (Q) for disease Y that is 20\% effective at reducing disease-specific mortality. Calculate the life expectancy of a 55-year-old woman with disease Y who is receiving treatment Q.

<span style="color:red">SOLUTION</span>


We begin by calculating a new disease-specific hazard rate under treatment (20\% reduction):

\[\mu_{y_{tx}}  = 0.8 \mu_Y\]

```{r}
mu.y.tx <- 0.8 * mu.y
print(mu.y.tx)
```

Now we can calculate Life Expectancy for a 55 year old woman with disease Y who is receiving treatment Q:

\[LE = \frac{1}{\mu} = \frac{1}{\mu_{ASR} + \mu_{y_{tx}}}\]

```{r}
LE.55 <- 1/(mu.asr.55 + mu.y.tx)
print(LE.55)
```

### Probability Estimation


(a) A trial that compared a cholesterol-lowering drug versus placebo estimated 5-year survival probabilities of 85\% for the placebo arm and 90\% for the treatment arm. The expected 5-year probability of dying for a cohort of similar age, sex, and race over 5 years is 3\%.


(i.) What is the percent reduction in the overall mortality rate provided by treatment? What is the percent reduction in the disease-specific mortality rate provided by treatment?


<span style="color:red">SOLUTION</span>

In the **placebo group**, the 5 year survival probability is 85\% , implying:

\[S(5) = 0.85 \implies 0.85 = e^{-5\mu_{placebo}} \implies \mu_{placebo} = \frac{ln(0.85)}{-5}\]

```{r}
mu.placebo <- log(0.85)/-5
print(mu.placebo)
```

In the **treatment group**, the 5 year survival probability is 90\% , implying:

\[S(5) = 0.9 \implies 0.9 = e^{-5\mu_{treat}} \implies \mu_{treat} = \frac{ln(0.9)}{-5}\]

```{r}
mu.treat <- log(0.9)/-5
print(mu.treat)
```


The percent risk reduction in **overall** mortality rate can be calculated:

\[ = (1 - \frac{\mu_{treat}}{\mu_{placebo}})*100\]

```{r}
risk.reduction.overall <- (1 - (mu.treat/mu.placebo))*100
print(risk.reduction.overall)
```

The background 5 year survival probability is 97\% , implying:

\[S(5) = 0.97 \implies 0.97 = e^{-5\mu_{ASR}} \implies \mu_{ASR} = \frac{ln(0.97)}{-5}\]

```{r}
mu.asr <- log(0.97)/-5
print(mu.asr)
```

We can calculate the disease specific mortalities, $\mu_{treat}$ and $\mu_{placebo}$:


\[\mu_{placebo}' = \mu_{placebo} - \mu_{ASR}\]


\[\mu_{treat}' = \mu_{treat} - \mu_{ASR}\]


```{r}
mu.placebo.prime <- mu.placebo - mu.asr
mu.treat.prime <- mu.treat - mu.asr

print(mu.placebo.prime)
print(mu.treat.prime)
```


Percent risk reduction in **disease-specific** mortality rate:

```{r}
risk.reduction.disease <- (1 - (mu.treat.prime/mu.placebo.prime))*100
print(risk.reduction.disease)
```




(ii.) The clinical trial also estimated 5-year “event-free” survival probabilities of 78% for the placebo arm and 85% for the treatment arm, where event is defined as either death from any cause or a nonfatal heart attack. What is the rate of nonfatal heart attack in each arm?


<span style="color:red">SOLUTION</span>


In the **placebo group**, the 5 year survival probability is 78\% , implying:

\[S(5) = 0.78 \implies 0.78 = e^{-5\mu_{placebo}} \implies \mu_{placebo} = \frac{ln(0.78)}{-5}\]

```{r}
mu.placebo.ef <- log(0.78)/-5
print(mu.placebo.ef)
```

In the **treatment group**, the 5 year survival probability is 85\% , implying:

\[S(5) = 0.85 \implies 0.85 = e^{-5\mu_{treat}} \implies \mu_{treat} = \frac{ln(0.85)}{-5}\]

```{r}
mu.treat.ef <- log(0.85)/-5
print(mu.treat.ef)
```


Rate of non-fatal heart attacks ($\mu_{NFMI}$) = total event rate - mortality rate:

\[\mu_{NFMI_{placebo}} = 0.0497 - 0.0325\]

```{r}
mu.placebo.ef - mu.placebo
```
\[\mu_{NFMI_{treat}} = 0.0325 - 0.0211\]


```{r}
mu.treat.ef - mu.treat
```

(b) The following table provides data on the cumulative probability of slipping on ice for persons moving to Alaska. You may assume no deaths occur during this time period.


| |  |
|---|---|
| Year ($t$)  | Pr (Event $< t$)  |
| 1  |  0.08 |
| 2  |  0.12 |
| 3  | 0.15  |
| 5  |  0.20 |

Calculate the annual rate of ice slipping assuming that it is constant over the five years. Now calculate annual rates for each time period (you will assume that the annual rates in the fourth and fifth years are the same). What might be a reasonable rate to assume for Year 6?







